---
output: 
  rmarkdown::html_vignette:
    keep_md: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteIndexEntry{typical use case}
  %\usepackage[UTF-8]{inputenc}
---

```{r, include = FALSE}
# do not execute on CRAN: 
# https://stackoverflow.com/questions/28961431/computationally-heavy-r-vignettes
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
#knitr::opts_chunk$set(eval = !is_check)
```

```{r setup, include = FALSE}
#rmarkdown::render("vignettes/useCase.Rmd")
#knitr::opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(
    #, fig.align = "center"
    #, fig.width = 3.27, fig.height = 2.5, dev.args = list(pointsize = 10)
    #,cache = TRUE
    #, fig.width = 4.3, fig.height = 3.2, dev.args = list(pointsize = 10)
    #, fig.width = 6.3, fig.height = 6.2, dev.args = list(pointsize = 10)
    # works with html but causes problems with latex
    #,out.extra = 'style = "display:block; margin: auto"' 
    )
knitr::knit_hooks$set(spar = function(before, options, envir) {
    if (before) {
        par(las = 1 )                   #also y axis labels horizontal
        par(mar = c(2.0,3.3,0,0) + 0.3 )  #margins
        par(tck = 0.02 )                          #axe-tick length inside plots             
        par(mgp = c(1.1,0.2,0) )  #positioning of axis title, axis labels, axis
     }
})
```


```{r, include = FALSE, warning = FALSE}
#themeTw <- theme_bw(base_size = 10) + theme(axis.title = element_text(size = 9))
#bgiDir <- "~/bgi"
```

# NetCDF in R Tutorial

from [Dept. Geography, Univ. Oregon](http://geog.uoregon.edu/GeogR/topics/netCDF-in-R.html)

## Reading a netCDF data set using the ncdf4 package

The ndcf4 package works much the same way as the ncdf package, but the functions in this package are a little more similar to the functions in the netCDF Fortran or C libraries than are those in the ncdf package. To begin, load the ncdf4 package.

```{r}
library(ncdf4)
```

Open the NetCDF data set, and print some basic information. The print() function applied to the ncin object produces information similar to that produced by the command-line utility ncdump.

```{r}
ncname <- "cru10min30_tmp"  
ncfname <- paste(ncname,".nc", sep = "")
dname <- "tmp"  # note: tmp means temperature (not temporary)
# open a netCDF file
getwd()
ncin <- nc_open(ncfname)
print(ncin)
```


Get the longtiudes and latitudes as before, using now the ncvar_get() function in ncdf4.
```{r}
lon <- ncvar_get(ncin,"lon")
nlon <- dim(lon)
head(lon)

lat <- ncvar_get(ncin,"lat",verbose = F)
nlat <- dim(lat)
head(lat)

print(c(nlon,nlat))
```

Get the time variable and its attributes using the ncvar_get() and ncatt_get() functions, and list them, and also get the number of times using the dim() function.
```{r}
t <- ncvar_get(ncin,"time")
t

tunits <- ncatt_get(ncin,"time","units")
nt <- dim(t)
nt
```


Print the time units string. Note the structure of the time units attribute. The object tunits has two components hasatt (a logical variable), and tunits$value, the actual “time since” string.

```{r}
tunits
```

Get the input variable (tmp) and its attributes, and verify the size of the array.

```{r}
tmp_array <- ncvar_get(ncin,dname)
dlname <- ncatt_get(ncin,dname,"long_name")
dunits <- ncatt_get(ncin,dname,"units")
fillvalue <- ncatt_get(ncin,dname,"_FillValue")
dim(tmp_array)
```


Get the global attributes.
```{r}
title <- ncatt_get(ncin,0,"title")
institution <- ncatt_get(ncin,0,"institution")
datasource <- ncatt_get(ncin,0,"source")
references <- ncatt_get(ncin,0,"references")
history <- ncatt_get(ncin,0,"history")
Conventions <- ncatt_get(ncin,0,"Conventions")
```


Close the netCDF file using the nc_close() function.

```{r}
nc_close(ncin)
```


Check what’s in the current workspace:
```{r}
ls()
outworkspace = "netCDF01.RData"
save.image(file = outworkspace)
```



